---
title: 'Using PCA to Identify the Orientation of Polygons'
author: "John Koo"
# output: pdf_document
output: html_document
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      dpi = 300, 
                      fig.align = 'center')
```

```{r example_image}
example.image <- png::readPNG('~/dev/polygon-labels/FDI102A.png')
grid::grid.raster(example.image)
```

# Motivating example

When generating maps consisting of multiple regions with defined boundaries 
(e.g., polygons), it's often times necessary to label the regions. A natural 
method of labeling might be to just identify the centroid of each region and 
place the labels (text) there. Depending on the specific shape of each region, 
this may not work as intended. It may be more appropriate to rotate the text to 
better fit the shape of the polygon. 

In the following example, a map of New England is generated, and each state is 
labeled.

```{r no_angle_example, fig.height = 4}
# packages used in this code:
#  * magrittr
#  * dplyr
#  * ggplot2
#  * import
#  * rgeos
#  * broom
#  * foreach

# load packages
import::from(magrittr, `%>%`, `%$%`, extract)
import::from(ggplot2, 
             ggplot, geom_polygon, geom_text, geom_point, geom_segment, aes, 
             coord_map, 
             map_data)
import::from(foreach, foreach, `%do%`)
import::from(broom, tidy)
dp <- loadNamespace('dplyr')
rg <- loadNamespace('rgeos')

#' @title Turn a data frame of vertices into a polygon/multipolygon
#' @param input.df (data frame) Data frame of vertices, can contain multiple 
#;   polygons
#' @param x.col (character)
#' @param y.col (character)
#' @param polygon.col (character)
#' @return (SpatialPolygons)
df.to.sp <- function(input.df, x.col = 'long', y.col = 'lat', polygon.col) {
  # separate out each polygon 
  polygon.ids <- unique(input.df[[polygon.col]])
  sapply(polygon.ids, function(pid) {
    # subset input.df to each polygon
    polygon.df <- input.df %>% 
        dplyr::filter_(paste0(polygon.col, ' =="', pid, '"'))
    # start compiling a WKT
      paste(polygon.df[[x.col]], polygon.df[[y.col]]) %>% 
        paste(collapse = ', ') %>% 
        paste0('((', ., '))')
  }) %>% 
    paste(collapse = ', ') %>% 
      paste0('MULTIPOLYGON(', ., ')') %>% 
      rgeos::readWKT()
}

# state boundary data
states.df <- map_data('state') %>% 
  dp$mutate(region = gsub(' ', '\n', region))
head(states.df)

# vector of new england states
ne.states <- c('massachusetts', 'connecticut', 'rhode\nisland', 
               'vermont', 'new\nhampshire', 'maine')

# data frame of new england states
ne.df <- dp$filter(states.df, region %in% ne.states)

# data frame of centroids
#  1. convert each state into a spatial object
#  2. use built-in spatial functions to find the centroid of each state
#  3. compile into a data frame
ne.labels.df <- foreach(ne.state = ne.states, .combine = dp$bind_rows) %do% {
  temp.sp <- df.to.sp(dp$filter(ne.df, region == ne.state), 
                      polygon.col = 'group')
  temp.centroid <- rg$gCentroid(temp.sp)
  data.frame(state = ne.state, 
             long = temp.centroid$x, 
             lat = temp.centroid$y, 
             stringsAsFactors = FALSE)
}

# plot
ggplot() + 
  coord_map() + 
  geom_polygon(data = ne.df, 
               colour = 'black', fill = 'white', 
               aes(x = long, y = lat, group = group)) + 
  geom_text(data = ne.labels.df, 
            aes(x = long, y = lat, label = state))
```

Looking at this plot, we might think that maybe the labels for New Hampshire 
and Vermont should be rotated 90 degrees. We can do this manually:

```{r manual_rotation}
ggplot() + 
  coord_map() + 
  geom_polygon(data = ne.df, 
               colour = 'black', fill = 'white', 
               aes(x = long, y = lat, group = group)) + 
  geom_text(data = ne.labels.df, 
            aes(x = long, y = lat, label = state, 
                angle = c(0, 0, 0, 90, 90, 0)))
```

(We can also see that perhaps the centroid might not be the best choice for the 
position of the label.)

Ideally, this would be done automatically/algorithmically. In order to come up 
with a method, we might want to think about the polygons as "pointing" in a 
certain direction. Vermont and New Hampshire point up and down, and 
Massachusetts and Connecticut point left to right. Maine points 45 degrees 
(although we might still want to have the label horizontal), and Rhode Island is 
relatively square/spherical. 

# Method

Let's look at one state that is clearly neither horizontal nor vertical. This 
time, instead of looking at the polygon, let's focus on the vertices.

```{r ca_plot}
ca.df <- dp$filter(states.df, region == 'california')
ca.plot <- ggplot(ca.df) + 
  geom_point(aes(x = long, y = lat), size = .1) + 
  coord_map()
ca.plot
```

If we think of the vertices as Euclidean data, we can say that there is some 
correlation. Then, similar to the Principal Component Analysis method, we can 
rotate our coordinates to reduce the correlation. 

```{r ca_rotate}
# pca on the vertices
ca.pca <- ca.df %>% 
  dp$select(long, lat) %>% 
  princomp()
ca.pca$loadings

# we also want the centroid
ca.sp <- df.to.sp(ca.df, polygon.col = 'group')
ca.centroid <- rg$gCentroid(ca.sp)

ca.plot + 
  geom_segment(aes(x = ca.centroid$x, 
                   xend = ca.centroid$x + ca.pca$loadings[1, 1] * ca.pca$sdev[1], 
                   y = ca.centroid$y, 
                   yend = ca.centroid$y + ca.pca$loadings[2, 1] * ca.pca$sdev[1]), 
               colour = 'red') + 
  geom_segment(aes(x = ca.centroid$x, 
                   xend = ca.centroid$x + ca.pca$loadings[1, 2] * ca.pca$sdev[2], 
                   y = ca.centroid$y, 
                   yend = ca.centroid$y + ca.pca$loadings[2, 2] * ca.pca$sdev[2]), 
               colour = 'red')
```

So we want the angle of the first principal component.

```{r ca_angle}
ca.angle <- atan2(ca.pca$loadings[2, 1], 
                  ca.pca$loadings[1, 1])
ca.angle
ca.angle.deg <- ca.angle * 180 / pi
ca.angle.deg

ggplot(ca.df) + 
  coord_map() + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               colour = 'black', fill = 'white') + 
  geom_text(aes(x = ca.centroid$x, y = ca.centroid$y, 
                label = 'california', angle = ca.angle.deg))
```

And to make things more readable:

```{r ca_angle_corrected}
ca.angle.deg <- ca.angle.deg + 180
ca.angle.deg

ggplot(ca.df) + 
  coord_map() + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               colour = 'black', fill = 'white') + 
  geom_text(aes(x = ca.centroid$x, y = ca.centroid$y, 
                label = 'california', angle = ca.angle.deg))
```

## Additional Consideration

This method can break down easily when the points are not uniformly spaced 
around the polygon and when the polygon is not convex. Although are true here, 
in this case the method still behaves well, but we can easily come up with 
examples where this breaks down. To alleviate some (but certainly not all) of 
this, we can simplify the polygon by looking at its convex hull (and this 
functionality is built into the `rgeos` package already, so we don't have to 
really do anything extra). 

So this time, we first take the convex hull of the original polygon, compile 
the new vertices into a data matrix, perform principal component analysis on 
the new vertices, and then compute the angle of the first component. 

```{r convex_hull}
ca.convex.hull.sp <- rg$gConvexHull(ca.sp)

# we don't want to change the centroid, but we want to recompute the angle
# so instead of using the original set of vertices, obtain a set of vertices 
# the convex hull (again, package functions)
ca.convex.hull.df <- tidy(ca.convex.hull.sp)
ca.convex.hull.pca <- ca.convex.hull.df %>% 
  dp$select(long, lat) %>% 
  princomp()
new.loadings <- ca.convex.hull.pca$loadings

new.angle <- atan2(new.loadings[2, 1], 
                   new.loadings[1, 1]) * 180 / pi + 180
new.angle

ggplot(ca.df) + 
  coord_map() + 
  geom_polygon(aes(x = long, y = lat, group = group), 
               colour = 'black', fill = 'white') + 
  geom_text(aes(x = ca.centroid$x, y = ca.centroid$y, 
                label = 'california', angle = new.angle))
```

In this case, there isn't much of a change since most of the original vertices 
lie along the western border, which follows the same general shape as the 
polygon.

# Implementation

```{r function}
#' @title Assign angles and centroids for polygon labels
#' @description Given a tidy data frame of polygons or multipolygons, find the 
#'   centroid and approximate orientation using principal component analysis to 
#'   automatically come up with a nice way to label the polygons on a map.
#' @param polygon.df (data frame) A tidy data frame containing (multi)poygon 
#'   vertices and labels
#' @param x.col (character)
#' @param y.col (character)
#' @param id.col (character)
#' @param subid.col (character)
#' @return (data frame)
polygon.labels <- function(polygon.df, 
                           x.col = 'long', y.col = 'lat', 
                           id.col, subid.col) {
  # find vector of (multi)polygon ids
  id.vec <- unique(polygon.df[[id.col]])
  
  foreach::foreach(id = id.vec, .combine = dplyr::bind_rows) %do% {
    # subset to just this polygon
    multipol.df <- polygon.df %>%
      dplyr::filter_(paste0(id.col, ' == "', id, '"'))
    
    # vector of sub-ids (for each piece)
    subid.vec <- unique(multipol.df[[subid.col]])
    
    # construct a spatial object for the id
    vertices.sp <- foreach::foreach(subid = subid.vec, .combine = c) %do% {
      pol.df <- multipol.df %>% 
        dplyr::filter_(paste0(subid.col, ' =="', subid, '"'))
      paste(pol.df[[x.col]], pol.df[[y.col]]) %>% 
        paste(collapse = ', ') %>% 
        paste0('((', ., '))')
    } %>% 
      paste(collapse = ', ') %>% 
      paste0('MULTIPOLYGON(', ., ')') %>% 
      rgeos::readWKT()
    
    # PCA on vertices to determine the centroid
    # note that we're starting with vertices.sp instead of multipol.df
    # so we can do some simplification first
    # then find the angle of the first component
    principal.angle <- vertices.sp %>% 
      # simplify a bit (WIP)
      rgeos::gConvexHull() %>% 
      # turn into long data frame
      broom::tidy() %>% 
      # only want spatial coordinates
      dplyr::select(long, lat) %>% 
      # remove repeats (e.g., opening and ending of polygon)
      dplyr::distinct() %>% 
      cov() %>% 
      eigen() %$%
      vectors %>% 
      extract(1:2) %>% 
      rev() %>% 
      as.list() %>% 
      do.call(atan2, .) %>% 
      {. * 180 / pi + 180}
    
    # adjust the angle for better readability
    # it should go from pi / 4 to 9 pi / 4
    if (principal.angle > 180 & principal.angle < 270) {
      principal.angle <- principal.angle + 180
    }
    
    # also need the centroid
    centroid <- vertices.sp %>% 
      rgeos::gCentroid()
    
    dplyr::data_frame(id, principal.angle, 
                      x = centroid$x, y = centroid$y)
  }
}

midwest.states <- c('indiana', 'ohio', 'illinois', 'wisconsin', 'michigan', 
                    'minnesota', 'iowa', 'missouri', 'kansas', 'nebraska', 
                    'north\ndakota', 'south\ndakota')

midwest.df <- dp$filter(states.df, region %in% midwest.states)
midwest.labels.df <- polygon.labels(midwest.df, 
                                    id.col = 'region', subid.col = 'group')

midwest.plot <- ggplot() + 
  geom_polygon(data = midwest.df, 
               aes(x = long, y = lat, group = group), 
               colour = 'black', fill = 'white') + 
  geom_text(data = midwest.labels.df, 
            aes(x = x, y = y, label = id, angle = principal.angle)) + 
  coord_map()
midwest.plot

ne.states <- c('maine', 'new\nhampshire', 'vermont', 
               'massachusetts', 'connecticut', 'rhode\nisland')

ne.df <- dp$filter(states.df, region %in% ne.states)
ne.labels.df <- polygon.labels(ne.df, id.col = 'region', subid.col = 'group')

ne.plot <- ggplot() + 
  geom_polygon(data = ne.df, aes(x = long, y = lat, group = group), 
               colour = 'black', fill = 'white') + 
  geom_text(data = ne.labels.df, 
            aes(x = x, y = y, label = id, angle = principal.angle)) + 
  coord_map()
ne.plot
```

Cases where this was unsuccessful (or not as successful as we would like):

* Large states for which a horizontal label would fit regardless of the shape or
orientation (e.g., Maine, Missouri)
* States with a "slight" angle (e.g., Massachusetts, North Dakota)---perhaps in 
these cases it would be better to "snap" to 0 degrees
* States for which the centroid is not the optimal location of the label (e.g., 
Michigan, Massachusetts)
* States where even the convex hull is not sufficiently simple (e.g., Indiana)
* States that are not sufficiently ellipsoid-shape (e.g., Minnesota, Missouri)